Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-08 paper-ieee-trans.tex
Introduction of section III rephrased

INDIGO--DataCloud is the most recent project included in this paper thus it
emerged from the lessons learned in the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years in terms
of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of metrics to measure the
success of the solutions implemented, has been complemented by the added 
complexity of being framed into large collaborative projects with an extensive
list of partners.

In the following section we will describe the three main pillars that sustain the project's strategy to seek reliability in the 
software produced. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} guides the development of the software in
a per-change basis, validating the software at the very early stages. The 
software delivery process is designed as a bottom-up \textit{adoption of DevOps
principles}, passing through Continuous Integration (CI) and Continuous
Delivery (CD) approaches. Once the software is released, it follows a 
\textit{two-factor validation} in the project's pilot infrastructures and
resource provider's early-adoption.

@@

\begin{itemize}
\item \textbf{Code style}.
Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standard} is initially recommended, however the developer teams eventually
choose the set of guidelines to comply with. The ultimate goal for code style assessment is to
improve the readability and reusability of the source code produced under the scope of the
project. Figure \ref{fig:fig_codestyle} showcases the code style standards and their popularity
over the software components contributing to the project.

\item \textbf{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major INDIGO--DataCloud releases, showing an increment in the 
averaged value of {\sl INDIGO--2} release. As it can be seen, a very small set of products with low coverage
values highly impacted the overall result, although a significant part of these products are upstream
contributions or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them had values above
the 50\% coverage.

\item \textbf{Integration testing}. Software components usually interact with other services while on
operation state. Integration testing deals with the interactions among coupled software components or
parts of a system that cooperates to achieve a given functionality. This type of testing might be 
complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
guarantee the overall operation of the component, focused on the services it intefaces with, whenever
new functionalities are introduced.

\item \textbf{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
production.

\item \textbf{Documentation}.
Last but not least, the SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. These requirements also 
promote the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and 
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

\item \textbf{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation, 
self-documenting code, as they sequentially guide the component's deployment process in multiple
platforms. In this regard, INDIGO--DataCloud project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

>>>>

INDIGO--DataCloud is the most recent project included in this paper thus it
emerged from the lessons learned in the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of
being framed into large collaborative projects with an extensive list of partners.

In the following section we will describe the three main pillars that sustain the project's strategy to seek reliability throughout the 
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} will provide guidelines to develop quality 
software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is constituted by a \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

\begin{itemize}
\item \textit{Code style}.
Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standard} is initially recommended, however the developer teams eventually
choose the set of guidelines to comply with. The ultimate goal for code style assessment is to
improve the readability and reusability of the source code produced under the scope of the
project. Figure \ref{fig:fig_codestyle} showcases the code style standards and their popularity
over the software components contributing to the project.

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major INDIGO--DataCloud releases, showing an increment in the 
averaged value of {\sl INDIGO--2} release. As it can be seen, a very small set of products with low coverage
values highly impacted the overall result, although a significant part of these products are upstream
contributions or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them had values above
the 50\% coverage.

\item \textit{Integration testing}. Software components usually interact with other services while on
operation state. Integration testing deals with the interactions among coupled software components or
parts of a system that cooperates to achieve a given functionality. This type of testing might be 
complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
guarantee the overall operation of the component, focused on the services it intefaces with, whenever
new functionalities are introduced.

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
production.

\item \textit{Documentation}.
Last but not least, the SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. These requirements also 
promote the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and 
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

\item \textit{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation, 
self-documenting code, as they sequentially guide the component's deployment process in multiple
platforms. In this regard, INDIGO--DataCloud project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}
